a.outliers

### 2022-09-20 hw4-1/ashish/a.outliers.Rmd
outlier ques final

| Variable | Description                                                                            |
|-------------------|-----------------------------------------------------|
| M        | percentage of males aged 14--24 in total state population                              |
| So       | indicator variable for a southern state                                                |
| Ed       | mean years of schooling of the population aged 25 years or over                        |
| Po1      | per capita expenditure on police protection in 1960                                    |
| Po2      | per capita expenditure on police protection in 1959                                    |
| LF       | labour force participation rate of civilian urban males in the age-group 14-24         |
| M.F      | number of males per 100 females                                                        |
| Pop      | state population in 1960 in hundred thousands                                          |
| NW       | percentage of nonwhites in the population                                              |
| U1       | unemployment rate of urban males 14--24                                                |
| U2       | unemployment rate of urban males 35--39                                                |
| Wealth   | wealth: median value of transferable assets or family income                           |
| Ineq     | income inequality: percentage of families earning below half the median income         |
| Prob     | probability of imprisonment: ratio of number of commitments to number of offenses      |
| Time     | average time in months served by offenders in state prisons before their first release |
| Crime    | crime rate: number of offenses per 100,000 population in 1960                          |

@@

```{r}
aggregate(cbind(Pop) ~ So, crime_df, sum)

#split by Southern states
south_split = aggregate(cbind(Ineq,M,M.F,U1,Time,NW,Ed) ~ So, crime_df, mean)
south_split
```

@@

```{r}
#Boxplot of crime
boxplot(crime_df$Crime, horizontal = TRUE)
```

```{r}
boxplot(crime_df$Crime, horizontal = TRUE)
```

```{r}
#Plot crime by population, expect this to be straight line
plot(x=crime_df$Pop,y=crime_df$Crime)
```

@@

The two salient points here are:

@@

Basis Histogram the data presents deviation from normality as the distribution is not symmetric, especially without the log transformation\

@@

Similarly from QQ plot, we see that the points diverge from line especially at the extremes, hinting at non-normality in that region. While log of crime is relatively more normal

@@

```{r}
#boxplot(crime_df$Crime ~ crime_df$So)
boxplot(crime_df$Crime)
```

```{r}
#boxplot(log(crime_df$Crime))
```

two points seem pretty far away from the whiskers.

@@

While the Crime variable is already normalised, number of offenses per 100,000 population in 1960. such high crime rate for such small population seems counter intuitive, especially so when all other relevant variables like Po1, Po2, Prob are fairly similar. Let us examine this in detail by plotting

```{r}
plot(x=crime_df$Pop,y=crime_df$Crime)
cor(crime_df$Pop,crime_df$Crime)
```

```{r}
cor(crime_df$Prob,crime_df$Crime)
```

@@

After removing 1993 point we get even lower p-value this time

```{r}
mask2 = crime_df$Crime<1960
grubbs.test(crime_df[mask2,]$Crime, type = 10, opposite = FALSE, two.sided = FALSE)
```

```{r}

```

>>>>

| Variable | Description                                                                            |
|----------|----------------------------------------------------------------------------------------|
| M        | percentage of males aged 14--24 in total state population                              |
| So       | indicator variable for a southern state                                                |
| Ed       | mean years of schooling of the population aged 25 years or over                        |
| Po1      | per capita expenditure on police protection in 1960                                    |
| Po2      | per capita expenditure on police protection in 1959                                    |
| LF       | labour force participation rate of civilian urban males in the age-group 14-24         |
| M.F      | number of males per 100 females                                                        |
| Pop      | state population in 1960 in hundred thousands                                          |
| NW       | percentage of nonwhites in the population                                              |
| U1       | unemployment rate of urban males 14--24                                                |
| U2       | unemployment rate of urban males 35--39                                                |
| Wealth   | wealth: median value of transferable assets or family income                           |
| Ineq     | income inequality: percentage of families earning below half the median income         |
| Prob     | probability of imprisonment: ratio of number of commitments to number of offenses      |
| Time     | average time in months served by offenders in state prisons before their first release |
| Crime    | crime rate: number of offenses per 100,000 population in 1960                          |

@@

### Difference b/w Southern and Northern states

```{r}
aggregate(Pop ~ So, crime_df, sum)
```

```{r}
south_split = aggregate(cbind(Crime,Ineq,M,M.F,U1,Time,NW,Ed) ~ So, crime_df, mean)
south_split
```

@@

```{r}
#Plot crime by population
plot(x=crime_df$Pop,y=crime_df$Crime)
```

The two extreme large points seem like potentail outliers

@@

The two salient assumptions here are:

@@

Basis Histogram the data presents deviation from normality as the distribution is not symmetric, especially without the log transformation. We can further test this with QQ plot

@@

Similarly from QQ plot, we see that the points diverge from line especially at the extremes, hinting at non-normality in that region. While log of crime is relatively more normal, the log transformation alters the scale of data, and we might want to detect outliers before applying it.

@@

```{r}
boxplot(crime_df$Crime)
```

As seen in scatter plot, there are two points seem pretty far away from the whiskers.

@@

While the Crime variable is already normalised, number of offenses per 100,000 population in 1960. such high crime rate for such small population seems counter intuitive, especially so when all other relevant variables like Po1, Po2, Prob are fairly similar. Let us examine this in detail by checking corelation:

```{r}
paste(cor(crime_df$Pop,crime_df$Crime),cor(crime_df$Prob,crime_df$Crime))
```

Basis population figures in 1960's, there are two candidate states for 1993 point: Wyoming & Nevada.

And comparing data from [Disatster Center](https://www.disastercenter.com/crime/wycrime.htm), Wyoming does have roughly twice the national average crime rate, which is similar to our data as well (1993/ 905.1). Maybe there was some particular event in 1960 in Wyoming, which is causing such behavior. To understand this better, we need some expertise of Crime rate Trends in Wyoming.

<https://en.wikipedia.org/wiki/1960_United_States_census#:~:text=The%20United%20States%20census%20of,enumerated%20during%20the%201950%20census.>

@@

After removing 1993 point we get even lower p-value this time, hinting that 1969 in absence of 1993 is a much stronger candidate for being a outlier.

Thus in conclusion, we have two candidate outliers in the data, 1993 and 1969 point. 1993 point is the highest crime rate and presents weak evidence to discard it as ouitlier, however, to be definitive, we need to study the Wyoming angle in detail, especially in relevance to exact modelling needs at hand.
